Reasonably Programmable Literal Notation

### 2018-02-20 icfp18/syntax-icfp18.tex
more progress on intro

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures. For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}. The example in Fig. \ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

@@

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable and persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.

@@

\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={context,free,syntax}]
  context-free syntax 
    (* ...direct mods to the CFG... *)
  (* ... *)
  let w = compute_w();
  let x = compute_x(w);
  let y = `(!R)@&{&/x!/:2_!x}'!R`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t by KQueryParser 
    using module Helpers as Helpers;
  (* ... *)
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the grammar fixed and making the segmentation and the type and binding structure explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\caption{Literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively uses the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}). 
The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:

\begin{enumerate}[leftmargin=12px]
\item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6? Might activating a new extension generate a conflicting expansion for the same form (i.e. might there be a syntactic ambiguity)?
\item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item[\Circle] \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
\item[\Circle] \textbf{(Context Dependence)} If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item[\Circle] \textbf{(Typing)} What type does \li{y} have? What type is each spliced expression expected to have?
\end{enumerate}

@@

An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against. With TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none]
  let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`
\end{lstlisting}
The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e. the characters between the outer delimiters, here backticks. Generalized literal forms syntactically subsume other literal forms because their bodies are left unparsed by the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
\begin{enumerate}[leftmargin=12px]
  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form. The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another.
  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no way to associate spliced terms with segments of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding.
  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed. Although no theorem about context independence was established, this appears to maintain context independence. However, this comes at a significant cost -- expansions cannot use any libraries whatsoever.
  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of the expected type. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml. These included non-trivial problems related to support for:
\begin{enumerate}[leftmargin=15px]
  \item \textbf{Structural Types}: There is no way to define literal notation for structural types, e.g. tuple and arrow types.
  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. for all types \li{list('a)}.
  \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g. for every abstract type defined by a module implementing the \li{QUEUE} signature.
  \item \textbf{ML-style Type Inference}: The type-directed dispatch mechanism relied on strictly local type inference, specified as a bidirectional type system \cite{Pierce:2000:LTI:345099.345100}. It is not clear that this is suitable for languages with ML-style constraint-based type inference.
  \item \textbf{Pattern Literals}: TSLs support only expression literals. No consideration was given to literals in pattern position, which are dual to expression literals in languages with support for ML-style pattern matching.
\end{enumerate}
\end{enumerate}
Taken together, we found that no existing approach for defining literal notation was suitable as-is for integration into Reason/OCaml.

\paragraph{Contributions} The aim of this paper is to introduce \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies. The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages. We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked \textbf{(a)-(e)} above. In describing these examples, we will also demonstrate that the literal notation itself can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems.

@@

The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are instead explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed. The applied TLM parses the literal body and generates an expansion statically. During the typing phase, a validation step ensures that programmers can reason abstractly about types, binding and segmentation. Taken together, the programmer can reason as follows:
\begin{enumerate}[nolistsep,leftmargin=10pt]
\item[\CIRCLE] \textbf{(Responsibility)} The applied TLM, \li{$kq}, is delegated responsibility for parsing the literal body. As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax that the TLM implements is unambiguous (even in situations where a spliced term itself applies a TLM, shown in a later example.)
\item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item[\CIRCLE] \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}. The library dependencies that the generated expansion might have are explicit (here, the expansion uses the module \li{Helpers}), and the system ensures that this dependency is available even if the identifier \li{Helpers} has been shadowed at the use site.
\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} determines the type that the expansion must have, here \li{KQuery.t}. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\end{enumerate}

>>>>

\vspace{-5px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures. For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}. The example in Fig. \ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

@@

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.

@@

\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={context,free,syntax}]
  context-free syntax 
    /* ...direct mods to the CFG... */
  /* ... */
  let w = compute_w();
  let x = compute_x(w);
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t by KQueryParser 
    using module Helpers as Helpers;
  /* ... */
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the grammar fixed and making the segmentation and the type and binding structure explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{3px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

\todo{show expr <- thing}Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively borrows the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}). 
The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:

\begin{enumerate}[nolistsep,noitemsep,leftmargin=12px]
\item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6? Might activating a new extension generate a conflicting expansion for the same form?
\item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item[\Circle] \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
\item[\Circle] \textbf{(Context Dependence)} If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item[\Circle] \textbf{(Typing)} What type does \li{y} have? What type is each spliced expression expected to have?
\end{enumerate}

@@

An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. Under this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against. So with TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none,basicstyle=\ttfamily\fontsize{9pt}{1em}\selectfont]
  let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`
\end{lstlisting}
The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e. the characters between the outer delimiters, here \li{`(} and \li{)`}, which are left unparsed by the fixed context-free syntax of the language (i.e. generalized literal forms are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
\begin{enumerate}[leftmargin=12px]
  \item[\CIRCLE] \textbf{(Responsibility)} The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined. The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another. 
  \item[\Circle] \textbf{(Segmentation)} The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e. subsequence) of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
  \item[\Circle] \textbf{(Capture)} The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)
  \item[\LEFTcircle] \textbf{(Context Independence)} The TSL mechanism as described requires that the expansion be completely closed, so it is trivially context independent. However, this comes at a significant expressive cost: the generated expansions cannot make use of any libraries whatsoever.
  \item[\LEFTcircle] \textbf{(Typing)} Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml \cite{ocaml-manual}:
\begin{enumerate}[leftmargin=15px]
  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types.
  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. at all types \li{list('a)}.
  \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g. at every abstract type defined by a module implementing the \li{QUEUE} signature. Parameterized and abstract types are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
  \item \textbf{ML-Style Type Inference}: The type-directed dispatch mechanism is specified by a bidirectional type system, which supports only local type inference \cite{Pierce:2000:LTI:345099.345100}. It is not clear that this is suitable for languages with ML-style constraint-based type inference.
  \item \textbf{Pattern Literals}: TSLs support only expression literals. No consideration was given to the problem of defining new literal notation in pattern position. Pattern matching is ubiquitous in ML-family languages. %Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
\end{enumerate}
\end{enumerate}
We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}. All together, we found that no existing approach was suitable as-is for integration into Reason/OCaml.

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies. The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages. We evaluate the first claim by developing a series of formal calculi equipped with proofs of these abstract reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked {(a)-(e)} above. In describing these examples, we also demonstrate that the literal parsers can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems for integrating new literal notation into a programming language.

@@

The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are instead explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed. The programmer can reason abstractly as follows:
\begin{enumerate}[leftmargin=12pt]
\item[\CIRCLE] \textbf{(Responsibility)} The parser named by applied TLM, here \li{KQueryParser} (Line 1), is exclusively responsible for parsing the literal body and generates an expansion statically. Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type. As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, shown in a later example.)
\item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item[\CIRCLE] \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item[\CIRCLE] \textbf{(Context Independence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}. The external dependencies of the generated expansion are explicit -- here, the expansion uses the module \li{Helpers} -- and the system ensures that this dependency is properly bound even if the identifier \li{Helpers} has been shadowed at the application site.
\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown). Moreover, each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference can be performed abstractly, i.e. given only the inferred segmentation rather than the full expansion\todo{say something formally about this?}. This segment-specific type information can also be communicated to the programmer upon request via the Merlin editor service. We will see examples of TLMs that make use of more advanced features of the OCaml type system in subsequent examples. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\end{enumerate}
